---
output:
  github_document:
    html_preview: FALSE
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%",
    dev = "ragg_png"
)
```

# geokit

<!-- badges: start -->
[![R-CMD-check](https://github.com/WangLabCSU/geokit/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/WangLabCSU/geokit/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/WangLabCSU/geokit/graph/badge.svg)](https://app.codecov.io/gh/WangLabCSU/geokit)
<!-- badges: end -->

The goal of `geokit` is to provide a comprehensive interface for most
interactions between R and [GEO database](https://www.ncbi.nlm.nih.gov/geo/).

## Installation

You can install the development version of `geokit` from
[R-universe](https://wanglabcsu.r-universe.dev) (Recommended):
```{r, eval=FALSE}
install.packages("geokit",
    repos = c(
        "https://wanglabcsu.r-universe.dev",
        "https://cloud.r-project.org"
    )
)
```

or from [GitHub](https://github.com/WangLabCSU/geokit) (Require `rust` to
compile) with:
```{r, eval=FALSE}
if (!requireNamespace("pak")) {
    install.packages("pak",
        repos = sprintf(
            "https://r-lib.github.io/p/pak/devel/%s/%s/%s",
            .Platform$pkgType, R.Version()$os, R.Version()$arch
        )
    )
}
pak::pak("WangLabCSU/geokit")
```

## Vignettes

```{r set_up}
library(geokit)
```

### Search GEO database
The NCBI uses a search term syntax which can be associated with a specific search field enclosed by a pair of square brackets. So, for instance `"Homo sapiens[ORGN]"` denotes a search for `Homo sapiens` in the `“Organism”` field. Details see <https://www.ncbi.nlm.nih.gov/geo/info/qqtutorial.html>. We can use the same term to query our desirable results in `geo_search`. `geo_search` will parse the searching results and return a `data.frame` object containing all the records based on the search term. The internal of `geo_search` is based on [`rentrez`](https://github.com/ropensci/rentrez) package, which provides functions working with the [NCBI Eutils](http://www.ncbi.nlm.nih.gov/books/NBK25500/) API, so we can utilize `NCBI API key` to increase the downloading speed, details see <https://docs.ropensci.org/rentrez/articles/rentrez_tutorial.html#rate-limiting-and-api-keys>.

Providing we want ***GSE*** GEO records related to ***human diabetes***, we can get these records by following code, the returned object is a `data.frame`:
```{r diabetes_gse_records, cache = TRUE}
diabetes_gse_records <- geo_search(
    "diabetes[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
)
head(diabetes_gse_records[1:5])
```

Once you have the search results, you can filter them based on specific
criteria. For instance, to filter for GSE datasets that contain at least 6
diabetic nephropathy samples with expression profiling, use the following code:

```{r filtered_diabetes_gse_records}
diabetes_nephropathy_gse_records <- diabetes_gse_records |>
    dplyr::mutate(
        number_of_samples = stringr::str_match(
            Contains, "(\\d+) Samples?"
        )[, 2L, drop = TRUE],
        number_of_samples = as.integer(number_of_samples)
    ) |>
    dplyr::filter(
        dplyr::if_any(
            c(Title, Summary),
            ~ stringr::str_detect(.x, "(?i)diabetes|diabetic")
        ),
        dplyr::if_any(
            c(Title, Summary),
            ~ stringr::str_detect(.x, "(?i)nephropathy")
        ),
        stringr::str_detect(Type, "(?i)expression profiling"),
        number_of_samples >= 6L
    )
head(diabetes_nephropathy_gse_records[1:5, 1:5])
```

After applying the filter, we obtain `r nrow(diabetes_nephropathy_gse_records)`
candidate datasets. This filtering step significantly reduces the time spent
manually reviewing summary records.

You can also use `geo_meta()` to dynamically create a self-knowledge-concerned
database in real-time.

### Download and Parse SOFT File from GEO database
The GEO database provides SOFT (Simple Omnibus Format in Text) files for GPL,
GSM, and GDS entities. SOFT is designed for efficient batch submission and
download of data. It is a simple, line-based, plain text format, which makes it
easy to generate SOFT files from common spreadsheet and database applications.

A single SOFT file can contain both data tables and accompanying descriptive
information for multiple, concatenated Platforms, Samples, and/or Series
records. The `geokit` package provides a `GEOSoft` class object to store SOFT
file contents. The `GEOSoft` object contains six slots: `accession`, `rcd_type`,
`rcd_name`, `metadata`, `datatable, and `columns`. 

 - `accession`: Stores the GEO accession ID.

 - `rcd_type`: Indicates the type of record (e.g., Platform, Sample, Series, Datasets). This helps categorize the data and identify the nature of the record.

 - `rcd_name`: Represents the name associated with the record (e.g., the GEO dataset name). It usually matches the accession, but in some cases, it may differ. This allows for a more flexible identification of the record.

 - `metadata`: Contains the header metadata from the SOFT file.

 - `datatable`: Contains the main data table, which is the primary data for
 analysis. 

 - `columns`: Provides descriptive column headers for the datatable.

You can use functions with the same names as the slots to extract the data.

The `geo_soft` function allows you to download and preprocess SOFT files. Here
are some example codes to download SOFT files for the GSM, GDS, and GPL
entities:
```{r gsm, cache = TRUE}
gsm <- geo_soft("GSM1", odir = tempdir())
gsm
head(datatable(gsm))
head(columns(gsm))
```

```{r gds, cache = TRUE}
gds <- geo_soft("GDS10", odir = tempdir())
gds
head(datatable(gds))
head(columns(gds))
```

For the GPL entity, the structure differs from that of GSM and GDS. The `geokit`
package provides the `GEOPlatform` class to store the contents of the GPL SOFT
file. A GPL SOFT file typically includes most of the contents found in its
subset entities, including both GSE and GSM. Therefore, the `GEOPlatform` class
contains both `gse` and `gsm` slots, each being a list of GEOSoft objects.

```{r gpl, cache = TRUE}
gpl <- geo_soft("gpl98", odir = tempdir())
gpl
```

Similarly, the GSE entity contains subset entities of GPL and GSM. The
`GEOSeries` class provides both `gpl` and `gsm` slots as lists of GEOSoft
objects.
```{r gse, cache = TRUE}
gse <- geo_soft("GSE10", odir = tempdir())
gse
```

### Download and Parse Seires matrix File from GEO database
In many typical analysis workflows, a series matrix file is commonly used. You
can easily handle it in `geokit` using the `gse_matrix()` function. The
`gse_matrix()` function returns an `ExpressionSet` object, which is compatible
with many Bioconductor packages.

```{r gse_matrix, cache = TRUE}
gse_matix <- geo_matrix("GSE180383", odir = tempdir())
gse_matix
```

When parsing `phenoData` from series matrix files, the `gse_matrix()` function
automatically discerns `characteristics_ch*` columns and parses multiple traits
from them. Each trait is named with the prefix `ch*`, corresponding to the
column name.

```{r}
Biobase::pData(gse_matix)[c("ch1_cultivar", "ch1_genotypes")]
```

By default, `gse_matrix()` attempts to map the GPL accession to a Bioconductor annotation package. You can control this behavior using the `add_gpl` parameter:

* Set `add_gpl = FALSE` to exclude feature information.
* Set `add_gpl = TRUE` to include platform information from GEO.

```{r}
Biobase::annotation(gse_matix)
```

### Download supplementary data from GEO database
GEO stores raw data and processed sequence data files as the external supplementary data files. Sometimes, we may want to preprocess and normalize the rawdata by ourselves, in addition, it's not uncommon that a GSE entity series matrix won't contain the expression matrix, which is almost the case of high-throughout sequencing data. `geo_suppl` is designed for these conditions. Usually, the expression matrix will be provided in the GSE supplementary files or in the GSM supplementary files.

If the expression matrix is given in the GSE supplementary files, we can download it directly use `geo_suppl`, which will return a character vector containing the path of downloaded files.
```{r gse160724, cache = TRUE}
gse160724 <- geo_suppl(
    "GSE160724",
    pattern = "counts_anno",
    odir = tempdir()
)
gse160724_dt <- data.table::fread(gse160724)
head(gse160724_dt[1:5])
```

### Other utilities
`geokit` also provide some useful function to help better interact with GEO.

- `geo_show` function: Require a geo entity id and open GEO Accession site in the default browser.
- `log_trans` function: Require a expression matrix and this function will check whether this expression matrix has experienced logarithmic transformation, if it hasn't, `log_trans` will do it. This is a helper function used in `GEO2R`. 

### sessionInfo
```{r}
sessionInfo()
```
