---
title: "Case Study: Urothelial Cancer Expression Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Urothelial Cancer Expression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a complete workflow for extracting, filtering, downloading, and analyzing urothelial cancer expression data from GEO. We'll cover dataset discovery, quality-based filtering, data retrieval, and exploratory analysis with publication-quality visualizations.

## Introduction

Urothelial carcinoma (bladder cancer) is one of the most common urological malignancies. Meta-analysis of gene expression studies can reveal consistent molecular alterations across patient cohorts. This workflow shows how to:

- Search and combine results for bladder/urothelial cancer datasets
- Filter datasets by sample size and platform type
- Download expression matrices
- Perform log transformation and data integration
- Conduct exploratory analysis (PCA, boxplots, heatmaps)
- Generate publication-quality visualizations

```{r setup, message = FALSE}
library(geokit)
library(dplyr)
library(stringr)
library(tidyr)
library(ggplot2)
```

## 1. Dataset Discovery

Search for bladder and urothelial cancer datasets using multiple related terms to maximize coverage.

```{r search_uc, cache = TRUE}
# Multiple search strategies for comprehensive coverage
uc_searches <- list(
  bladder = geo_search(
    "bladder cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
  ),
  urothelial = geo_search(
    "urothelial cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
  )
)

# Combine and deduplicate
uc_gse <- unique(dplyr::bind_rows(uc_searches))

cat(sprintf("Found %d unique urothelial cancer datasets\n", nrow(uc_gse)))
head(uc_gse[, 1:3])
```

## 2. Extract and Process Sample Information

```{r extract_info}
# Extract sample counts
uc_gse <- uc_gse |>
  dplyr::mutate(
    number_of_samples = stringr::str_match(
      Contains, "(\\d+) Samples?"
    )[, 2L, drop = TRUE],
    number_of_samples = as.integer(number_of_samples),
    # Extract platform information
    platform_count = stringr::str_match(
      Contains, "(\\d+) Platforms?"
    )[, 2L, drop = TRUE],
    platform_count = as.integer(platform_count)
  )

# Summary statistics
cat("\nSample size summary:\n")
summary(uc_gse$number_of_samples)
```

## 3. Filter Datasets by Quality Criteria

Apply stringent filters to select high-quality datasets suitable for meta-analysis.

```{r filter_quality}
uc_gse_filtered <- uc_gse |>
  dplyr::filter(
    # At least 20 samples for robust analysis
    number_of_samples >= 20,
    # Focus on expression profiling
    stringr::str_detect(Type, "(?i)expression profiling"),
    # Prefer microarray studies (exclude RNA-seq for now)
    stringr::str_detect(Type, "(?i)array"),
    # Exclude methylation studies
    !stringr::str_detect(Type, "(?i)methylation")
  ) |>
  dplyr::arrange(desc(number_of_samples))

cat(sprintf(
  "\nFiltered to %d high-quality datasets (%.1f%% of original)\n",
  nrow(uc_gse_filtered),
  100 * nrow(uc_gse_filtered) / nrow(uc_gse)
))

# Display top candidates
head(uc_gse_filtered[, c("Series Accession", "Title", "number_of_samples")], 10)
```

## 4. Download Expression Data

For demonstration, we'll work with a subset of datasets. In practice, you would download all filtered datasets.

```{r select_demo}
# Select top 3 datasets for demonstration
demo_gse <- head(uc_gse_filtered[["Series Accession"]], 3)
cat("Demo datasets:", paste(demo_gse, collapse = ", "), "\n")
```

```{r download_data, eval = FALSE}
# Download expression matrices
# Note: Using tempdir() for demo. For persistent storage, use a dedicated directory:
# odir <- "uc_expression_data"
# dir.create(odir, showWarnings = FALSE, recursive = TRUE)
expression_sets <- lapply(demo_gse, function(gse_id) {
  cat(sprintf("Downloading %s...\n", gse_id))
  tryCatch(
    {
      geo_matrix(gse_id, odir = tempdir())
    },
    error = function(e) {
      message(sprintf("Failed to download %s: %s", gse_id, e$message))
      return(NULL)
    }
  )
})

# Remove NULL entries (failed downloads)
expression_sets <- expression_sets[!sapply(expression_sets, is.null)]
names(expression_sets) <- sapply(expression_sets, function(x) {
  Biobase::annotation(x)
})

cat(sprintf("Successfully downloaded %d datasets\n", length(expression_sets)))
```

## 5. Data Transformation

Apply log2 transformation to ensure data comparability across datasets.

```{r transform_data, eval = FALSE}
# Apply log transformation
expression_sets_log <- lapply(expression_sets, function(eset) {
  log_trans(eset)
})

# Verify transformation
sapply(expression_sets_log, function(eset) {
  exprs <- Biobase::exprs(eset)
  c(
    min = min(exprs, na.rm = TRUE),
    median = median(exprs, na.rm = TRUE),
    max = max(exprs, na.rm = TRUE)
  )
})
```

## 6. Exploratory Analysis

### 6.1 Expression Distribution

Examine the distribution of expression values across samples using boxplots.

```{r expr_boxplot, eval = FALSE, fig.width = 10, fig.height = 6}
# Prepare data for plotting (first dataset as example)
eset <- expression_sets_log[[1]]
expr_matrix <- Biobase::exprs(eset)

# Convert to long format for ggplot
expr_df <- as.data.frame(expr_matrix) |>
  tibble::rownames_to_column("gene") |>
  tidyr::pivot_longer(
    cols = -gene,
    names_to = "sample",
    values_to = "expression"
  )

# Create boxplot
ggplot(expr_df, aes(x = sample, y = expression)) +
  geom_boxplot(fill = "lightblue", outlier.size = 0.5, outlier.alpha = 0.3) +
  labs(
    title = sprintf("Expression Distribution - %s", names(expression_sets_log)[1]),
    x = "Sample",
    y = "Log2 Expression"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major.x = element_blank()
  )
```

### 6.2 Principal Component Analysis (PCA)

Perform PCA to visualize sample relationships and identify potential batch effects.

```{r pca_analysis, eval = FALSE, fig.width = 8, fig.height = 6}
# Perform PCA on first dataset
eset <- expression_sets_log[[1]]
expr_matrix <- Biobase::exprs(eset)

# Remove genes with missing values
expr_matrix <- expr_matrix[complete.cases(expr_matrix), ]

# Perform PCA
pca_result <- stats::prcomp(t(expr_matrix), scale. = TRUE)

# Extract PC scores
pca_df <- as.data.frame(pca_result$x[, 1:2])
pca_df$sample <- rownames(pca_df)

# Calculate variance explained
var_explained <- summary(pca_result)$importance[2, 1:2] * 100

# Create PCA plot
ggplot(pca_df, aes(x = PC1, y = PC2)) +
  geom_point(size = 3, alpha = 0.7, color = "darkred") +
  geom_text(aes(label = sample), size = 2.5, vjust = -1) +
  labs(
    title = sprintf("PCA - %s", names(expression_sets_log)[1]),
    x = sprintf("PC1 (%.1f%% variance)", var_explained[1]),
    y = sprintf("PC2 (%.1f%% variance)", var_explained[2])
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid = element_line(color = "gray90")
  )
```

### 6.3 Sample Correlation Heatmap

Calculate and visualize pairwise sample correlations to assess data quality and identify outliers.

```{r correlation_heatmap, eval = FALSE, fig.width = 10, fig.height = 9}
# Calculate sample correlations
eset <- expression_sets_log[[1]]
expr_matrix <- Biobase::exprs(eset)
expr_matrix <- expr_matrix[complete.cases(expr_matrix), ]

# Compute correlation matrix
cor_matrix <- cor(expr_matrix, method = "pearson")

# Convert to long format for ggplot
cor_df <- as.data.frame(as.table(cor_matrix))
colnames(cor_df) <- c("Sample1", "Sample2", "Correlation")

# Create heatmap
ggplot(cor_df, aes(x = Sample1, y = Sample2, fill = Correlation)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0.8,
    limits = c(min(cor_df$Correlation), 1),
    name = "Pearson\nCorrelation"
  ) +
  labs(
    title = sprintf("Sample Correlation Heatmap - %s", names(expression_sets_log)[1]),
    x = "Sample",
    y = "Sample"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid = element_blank()
  )
```

### 6.4 Summary Statistics

Generate comprehensive summary statistics across datasets.

```{r summary_stats, eval = FALSE}
# Calculate summary statistics for each dataset
summary_stats <- lapply(names(expression_sets_log), function(name) {
  eset <- expression_sets_log[[name]]
  expr_matrix <- Biobase::exprs(eset)
  
  data.frame(
    dataset = name,
    n_samples = ncol(expr_matrix),
    n_genes = nrow(expr_matrix),
    mean_expr = mean(expr_matrix, na.rm = TRUE),
    median_expr = median(expr_matrix, na.rm = TRUE),
    sd_expr = sd(as.vector(expr_matrix), na.rm = TRUE),
    missing_pct = mean(is.na(expr_matrix)) * 100
  )
})

summary_table <- dplyr::bind_rows(summary_stats)
print(summary_table)

# Export summary
write.csv(summary_table, "uc_expression_summary.csv", row.names = FALSE)
```

## 7. Data Integration Considerations

When integrating multiple datasets for meta-analysis:

```{r integration_tips, eval = FALSE}
# 1. Check for common genes across datasets
common_genes <- Reduce(intersect, lapply(expression_sets_log, function(eset) {
  rownames(Biobase::exprs(eset))
}))

cat(sprintf("Common genes across all datasets: %d\n", length(common_genes)))

# 2. Subset to common genes
expression_sets_common <- lapply(expression_sets_log, function(eset) {
  eset[common_genes, ]
})

# 3. Combine expression matrices
combined_expr <- do.call(cbind, lapply(expression_sets_common, Biobase::exprs))

cat(sprintf(
  "Combined matrix dimensions: %d genes x %d samples\n",
  nrow(combined_expr),
  ncol(combined_expr)
))

# 4. Create study labels for batch correction
study_labels <- unlist(lapply(seq_along(expression_sets_common), function(i) {
  rep(names(expression_sets_common)[i], ncol(Biobase::exprs(expression_sets_common[[i]])))
}))

# Note: At this point, you would apply batch correction methods
# such as ComBat (sva package) or limma's removeBatchEffect
```

## 8. Visualization: Cross-Dataset Comparison

Compare expression patterns across multiple datasets.

```{r cross_dataset, eval = FALSE, fig.width = 10, fig.height = 6}
# Calculate mean expression per dataset
mean_expr_per_dataset <- sapply(expression_sets_log, function(eset) {
  rowMeans(Biobase::exprs(eset), na.rm = TRUE)
})

# Convert to data frame for plotting
mean_expr_df <- as.data.frame(mean_expr_per_dataset) |>
  tibble::rownames_to_column("gene") |>
  tidyr::pivot_longer(
    cols = -gene,
    names_to = "dataset",
    values_to = "mean_expression"
  )

# Create density plot
ggplot(mean_expr_df, aes(x = mean_expression, color = dataset)) +
  geom_density(linewidth = 1, alpha = 0.7) +
  labs(
    title = "Mean Expression Distribution Across Datasets",
    x = "Mean Log2 Expression",
    y = "Density",
    color = "Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )
```

## 9. Next Steps

With the downloaded and processed data, you can proceed to:

1. **Batch correction**: Use ComBat or similar methods to remove technical variation
2. **Differential expression**: Identify genes altered in cancer vs. normal samples
3. **Meta-analysis**: Combine results across studies using fixed/random effects models
4. **Pathway analysis**: Investigate enriched biological processes and pathways
5. **Validation**: Cross-validate findings in independent cohorts

```{r next_steps, eval = FALSE}
# Example: Extract phenotype data for differential expression
eset <- expression_sets_log[[1]]
pdata <- Biobase::pData(eset)

# View available phenotype columns
colnames(pdata)

# Typical workflow would identify tumor vs. normal samples
# and perform differential expression analysis
```

## Summary

This workflow demonstrated:

- Comprehensive dataset discovery using multiple search terms
- Filtering by sample size (≥20), platform type (microarray), and relevance
- Downloading expression matrices with `geo_matrix()`
- Log2 transformation using `log_trans()`
- Exploratory analysis including:
  - Expression distribution boxplots
  - Principal Component Analysis
  - Sample correlation heatmaps
  - Summary statistics
- Data integration considerations for meta-analysis

These steps provide a solid foundation for conducting rigorous gene expression meta-analysis in urothelial cancer research.

## Session Information
```{r}
sessionInfo()
```
