---
title: "Case Study: Diabetic Nephropathy Meta-Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Study: Diabetic Nephropathy Meta-Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette demonstrates a comprehensive meta-analysis workflow for diabetic nephropathy (DN) gene expression datasets. We'll show how to systematically search, filter, annotate, and visualize relevant GEO datasets to prepare for downstream meta-analysis.

## Introduction

Diabetic nephropathy is a major complication of diabetes and a leading cause of chronic kidney disease. Conducting a meta-analysis of gene expression studies can help identify robust molecular signatures. This workflow demonstrates how to use `geokit` to:

- Search GEO using multiple query strategies
- Build a customized metadata database
- Filter datasets by quality criteria
- Visualize dataset characteristics
- Prepare a curated list for downstream analysis

```{r setup}
library(geokit)
library(dplyr)
library(stringr)
library(ggplot2)
```

## 1. Multi-Strategy Dataset Discovery

We'll use multiple search terms to comprehensively identify diabetic nephropathy datasets. Different researchers may use varying terminology, so we query for both "diabetic nephropathy" and "diabetic kidney disease".

```{r search_dn, cache = TRUE}
# Define multiple search strategies
dn_search_terms <- c(
  "diabetic nephropathy[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]",
  "diabetic kidney disease[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
)

# Execute searches and combine results
dn_gse_list <- lapply(dn_search_terms, geo_search)
dn_gse <- unique(dplyr::bind_rows(dn_gse_list))

# Display summary
cat(sprintf("Found %d unique GSE datasets\n", nrow(dn_gse)))
head(dn_gse[, 1:3])
```

## 2. Extract Sample Information

Extract the number of samples from the "Contains" field to enable filtering by sample size.

```{r extract_samples}
dn_gse <- dn_gse |>
  dplyr::mutate(
    number_of_samples = stringr::str_match(
      Contains, "(\\d+) Samples?"
    )[, 2L, drop = TRUE],
    number_of_samples = as.integer(number_of_samples)
  )

# Quick statistics
summary(dn_gse$number_of_samples)
```

## 3. Filter by Quality Criteria

Apply filters for sample size, platform type, and study type to focus on high-quality expression profiling studies.

```{r filter_datasets}
dn_gse_filtered <- dn_gse |>
  dplyr::filter(
    # At least 6 samples for meaningful analysis
    number_of_samples >= 6,
    # Focus on expression profiling studies
    stringr::str_detect(Type, "(?i)expression profiling"),
    # Exclude certain platform types if needed
    !stringr::str_detect(Type, "(?i)methylation")
  )

cat(sprintf("After filtering: %d datasets (%.1f%% of original)\n",
            nrow(dn_gse_filtered),
            100 * nrow(dn_gse_filtered) / nrow(dn_gse)))
```

## 4. Build Metadata Database

Fetch detailed metadata for filtered datasets using `geo_meta()`. This creates a local database for offline analysis and detailed inspection.

```{r build_metadb, eval = FALSE}
# Create output directory for metadata
dn_metadb_dir <- "dn_metadb"
dir.create(dn_metadb_dir, showWarnings = FALSE, recursive = TRUE)

# Fetch metadata (this may take several minutes)
dn_metadb <- geo_meta(
  dn_gse_filtered[["Series Accession"]],
  odir = dn_metadb_dir
)

# Save for future use
saveRDS(dn_metadb, file.path(dn_metadb_dir, "dn_metadb.rds"))
```

```{r load_metadb, eval = FALSE}
# Load previously saved metadb
dn_metadb <- readRDS(file.path(dn_metadb_dir, "dn_metadb.rds"))

# Calculate actual sample counts from metadb
dn_metadb <- dn_metadb |>
  dplyr::mutate(
    number_of_samples = lengths(
      strsplit(Series_sample_id, "; ", fixed = TRUE)
    )
  )
```

## 5. Further Categorization

Categorize datasets by specific criteria relevant to diabetic nephropathy research.

```{r categorize, eval = FALSE}
# Identify datasets with specific DN-related keywords
dn_metadb <- dn_metadb |>
  dplyr::mutate(
    # Check for nephropathy mentions
    has_nephropathy = dplyr::if_any(
      c(Series_title, Series_summary, Series_overall_design),
      ~ stringr::str_detect(.x, "(?i)nephropath")
    ),
    # Check for kidney mentions
    has_kidney = dplyr::if_any(
      c(Series_title, Series_summary, Series_overall_design),
      ~ stringr::str_detect(.x, "(?i)kidney|renal")
    ),
    # Check for biopsy samples
    has_biopsy = dplyr::if_any(
      c(Series_title, Series_summary, Series_overall_design),
      ~ stringr::str_detect(.x, "(?i)biopsy|biopsies")
    ),
    # Extract tissue type mentions
    tissue_type = dplyr::case_when(
      stringr::str_detect(Series_title, "(?i)glomerul") ~ "glomerular",
      stringr::str_detect(Series_title, "(?i)tubul") ~ "tubular",
      stringr::str_detect(Series_title, "(?i)kidney|renal") ~ "kidney",
      TRUE ~ "other"
    )
  )

# Summary by tissue type
table(dn_metadb$tissue_type)
```

## 6. Visualization: Dataset Characteristics

### Timeline of Dataset Submissions

```{r timeline, eval = FALSE, fig.width = 8, fig.height = 4}
# Extract submission year from GEO accession or date fields
# Note: This assumes availability of date information in metadata
# You may need to adjust based on actual metadata structure
dn_metadb <- dn_metadb |>
  dplyr::mutate(
    submission_year = as.integer(
      stringr::str_extract(Series_geo_accession, "\\d{4}$")
    )
  )

# Create timeline plot
ggplot(dn_metadb, aes(x = submission_year)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(
    title = "Timeline of Diabetic Nephropathy GEO Dataset Submissions",
    x = "Submission Year",
    y = "Number of Datasets"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.minor = element_blank()
  )
```

### Sample Size Distribution

```{r sample_distribution, eval = FALSE, fig.width = 8, fig.height = 4}
ggplot(dn_metadb, aes(x = number_of_samples)) +
  geom_histogram(binwidth = 5, fill = "coral", color = "white") +
  geom_vline(
    xintercept = median(dn_metadb$number_of_samples, na.rm = TRUE),
    linetype = "dashed", color = "red", linewidth = 1
  ) +
  labs(
    title = "Distribution of Sample Sizes in DN Datasets",
    subtitle = sprintf(
      "Median: %d samples",
      median(dn_metadb$number_of_samples, na.rm = TRUE)
    ),
    x = "Number of Samples",
    y = "Number of Datasets"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    panel.grid.minor = element_blank()
  )
```

### Platform Usage Trends

```{r platform_trends, eval = FALSE, fig.width = 10, fig.height = 5}
# Extract platform information
platform_summary <- dn_metadb |>
  dplyr::count(Series_platform_id, sort = TRUE) |>
  dplyr::slice_max(n, n = 10)

ggplot(platform_summary, aes(x = reorder(Series_platform_id, n), y = n)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  coord_flip() +
  labs(
    title = "Top 10 Platforms Used in DN Studies",
    x = "Platform ID",
    y = "Number of Studies"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

## 7. Prepare Curated Dataset List

Create a final curated list of datasets suitable for meta-analysis.

```{r curate_list, eval = FALSE}
# Select high-quality datasets with sufficient samples
dn_curated <- dn_metadb |>
  dplyr::filter(
    number_of_samples >= 10,
    has_nephropathy | has_kidney
  ) |>
  dplyr::arrange(desc(number_of_samples)) |>
  dplyr::select(
    Series_geo_accession,
    Series_title,
    number_of_samples,
    Series_platform_id,
    tissue_type
  )

# Display top candidates
head(dn_curated, 10)

# Export for downstream analysis
write.csv(dn_curated, "dn_curated_datasets.csv", row.names = FALSE)
```

## 8. Next Steps for Meta-Analysis

With this curated list, you can proceed to:

1. **Download expression data**: Use `geo_matrix()` to download series matrix files
2. **Quality control**: Examine sample annotations and expression distributions
3. **Data integration**: Normalize and batch-correct across studies
4. **Differential expression**: Identify genes consistently dysregulated in DN
5. **Pathway analysis**: Investigate enriched biological processes

```{r download_example, eval = FALSE}
# Example: Download top dataset
top_gse <- dn_curated$Series_geo_accession[1]
eset <- geo_matrix(top_gse, odir = tempdir())

# Quick inspection
print(eset)
dim(Biobase::exprs(eset))
```

## Summary

This workflow demonstrated how to:

- Use multiple search strategies to comprehensively identify relevant datasets
- Build a customized metadata database with `geo_meta()`
- Apply quality filters for sample size and study type
- Visualize dataset characteristics over time
- Prepare a curated list for downstream meta-analysis

The systematic approach ensures reproducibility and helps identify the most suitable datasets for gene expression meta-analysis in diabetic nephropathy research.

## Session Information
```{r}
sessionInfo()
```
