---
title: "AI-Powered GEO Metadata Exploration with geo_chat()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AI-Powered GEO Metadata Exploration with geo_chat()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>",
eval = FALSE
)
```

# Introduction

The `geo_chat()` function enables interactive, AI-powered exploration of GEO
(Gene Expression Omnibus) metadata. By leveraging the
[ellmer](https://ellmer.tidyverse.org/) package, you can use natural language
to query, filter, and understand your GEO datasets.

> Warning: Be aware of the instability in results caused by some issues within the large language model itself.

## Prerequisites

Before using `geo_chat()`, you need to:

1. Install the `ellmer` package:
```{r install-ellmer}
install.packages("ellmer")
```

2. Configure an API key for your preferred AI provider. For example, for OpenAI:
```{r set-api-key}
# Set in your .Renviron file or directly:
Sys.setenv(OPENAI_API_KEY = "your-api-key-here")
```

Supported providers include:
- **OpenAI** (`ellmer::chat_openai`): Set `OPENAI_API_KEY`
- **Anthropic Claude** (`ellmer::chat_anthropic`): Set `ANTHROPIC_API_KEY`
- **Google Gemini** (`ellmer::chat_google_gemini`): Set `GOOGLE_API_KEY`
- And many more! See the [ellmer documentation](https://ellmer.tidyverse.org/) for the full list.

# Basic Usage

## Step 1: Get GEO Metadata

First, use geokit functions to retrieve metadata:

```{r setup}
library(geokit)

# Search for GEO Series related to your research topic
records <- geo_search("single cell RNA-seq[ALL] AND lung cancer[ALL] AND GSE[ETYP]")

# Get detailed metadata for the top results
meta <- geo_meta(records$`Series Accession`[1:10], odir = tempdir())
```

## Step 2: Start an AI Chat Session

Create a chat session with your metadata:

```{r create-chat}
chat <- geo_chat(meta, provider = ellmer::chat_openai)
```

Or OpenAI-compatible model:

```{r create-chat2}
chat <- geo_chat(meta,
                 provider = ellmer::chat_openai_compatible,
                 base_url = "https://open.cherryin.cc/v1",
                 model = "openai/gpt-4.1-mini", 
                 system_prompt = "Be terse")
```


## Step 3: Ask Questions!

Now you can interact with your metadata using natural language:
```{r chat-examples}
# Ask about the datasets
chat$chat("What types of studies are in this metadata?")

# Filter studies
chat$chat("Which studies have more than 50 samples?")

# Get R code suggestions
chat$chat("Write R code to filter studies that mention 'tumor' in the summary")

# Understand specific datasets
chat$chat("Tell me more about the study with the most samples")
```

Example output:

~~~
> chat$chat("What types of studies are in this metadata?")
The types of studies (Series_type) in this metadata are:

1. Expression profiling by high throughput sequencing
2. Expression profiling by high throughput sequencing; Genome variation profiling by high 
throughput sequencing
3. Expression profiling by high throughput sequencing; Other
4. Non-coding RNA profiling by high throughput sequencing
> chat$chat("Which studies have more than 50 samples?")
The study with more than 50 samples is:

- Series_geo_accession: GSE266330  
  - Series_sample_id count: 66 samples (GSM8245005 to GSM8245057 and GSM8669376 to GSM8669385)
> chat$chat("Write R code to filter studies that mention 'tumor' in the summary")
```r
# Assuming the data frame is named geo_data

tumor_studies <- subset(geo_data, grepl("tumor", Series_summary, ignore.case = TRUE))
```
> chat$chat("Tell me more about the study with the most samples")
The study with the most samples is Series_geo_accession GSE266330.

- Title: Single cell profiling of bone metastasis ecosystems from multiple cancer types reveals 
convergent and divergent mechanisms of bone colonization
- Summary: This study performed single-cell RNA-seq on 34 bone metastases (BMs) samples from 6 
cancer types to identify 3 ecosystem archetypes characterized by macrophages/osteoclasts, 
regulatory/exhausted T cells, and monocytes. Breast cancer BMs mostly exhibit 
macrophage-osteoclast archetype driven by osteolytic vicious cycle, kidney cancer BMs mainly 
belong to the regulatory/exhausted T cell archetype without osteoclasts, and lung cancer BMs are 
evenly distributed across all archetypes. The work reveals parallel immunosuppression and bone 
remodeling mechanisms and estrogen signaling differences in macrophages of the 
macrophage-osteoclast subtype.
- Overall design: Patient samples collected from multiple institutions, including 
surgery-obtained bone metastasis tissues from human patients and healthy bone marrow from 
individuals. Single-cell sequencing and pathology tissue sections were generated as relevant to 
clinical needs.
- Number of samples: 66 samples (GSM8245005 to GSM8245057 and GSM8669376 to GSM8669385)
- Contributor contact: Xiang Zhang (xiangz@bcm.edu), Baylor College of Medicine, Houston, USA
- Publication PubMed ID: 40412393
~~~


# Interactive Console Mode

For a more immersive experience, use the interactive console:

```{r live-console}
ellmer::live_console(chat)
```

This opens an interactive chat interface where you can have a back-and-forth
conversation with the AI about your GEO metadata.

# Using Different AI Providers

## OpenAI (GPT-4, GPT-3.5)

```{r openai}
chat <- geo_chat(meta, provider = ellmer::chat_openai, model = "gpt-4o")
```

## Anthropic Claude

```{r anthropic}
chat <- geo_chat(meta, provider = ellmer::chat_anthropic)
```

## Google Gemini

```{r gemini}
chat <- geo_chat(meta, provider = ellmer::chat_google_gemini)
```

# Customizing the AI Behavior

You can add custom instructions to guide the AI:

```{r custom-prompt}
chat <- geo_chat(
  meta,
  provider = ellmer::chat_openai,
  system_prompt = "
    Focus on identifying single-cell RNA-seq studies.
    When suggesting R code, use tidyverse syntax.
    Always explain your reasoning step by step.
  "
)
```

# Controlling Context Size

For large metadata tables, you can control how much data is sent to the AI:

```{r max-rows}
# Only include first 50 rows in context
chat <- geo_chat(meta, provider = ellmer::chat_openai, max_rows = 50)

# Disable column statistics summary
chat <- geo_chat(meta, provider = ellmer::chat_openai, include_summary = FALSE)
```

# Example Workflow: Finding Single-Cell Cancer Studies

Here's a complete workflow for finding relevant single-cell cancer studies:

```{r workflow}
library(geokit)

# 1. Search for relevant studies
cancer_gse <- geo_search(
  "single cell[ALL] AND cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
)
cat("Found", nrow(cancer_gse), "potential studies\n")

# 2. Get detailed metadata
meta <- geo_meta(
  cancer_gse$`Series Accession`[1:20],
  odir = "geo_cache"
)

# 3. Start AI chat session
chat <- geo_chat(meta, 
                 provider = ellmer::chat_openai_compatible,
                 base_url = "https://open.cherryin.cc/v1",
                 model = "openai/gpt-4.1-mini", 
                 system_prompt = "Be terse")

# 4. Use AI to analyze and filter
chat$chat("
  Identify studies that:
  1. Use scRNA-seq
  2. Include treatment and control samples
  
  List the accession numbers that match these criteria.
")

# 5. Get code to extract the filtered results
chat$chat("
  Write R code using dplyr to filter the metadata `meta`
  for the studies you identified.
")

```

Example output:

~~~
> # 1. Search for relevant studies
> cancer_gse <- geo_search(
+   "single cell[ALL] AND cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
+ )
→ Parsing GEO records                                         
Get records from NCBI for 4527 queries in 32.1s              
> cat("Found", nrow(cancer_gse), "potential studies\n")
Found 4527 potential studies
> # 2. Get detailed metadata
> meta <- geo_meta(
+   cancer_gse$`Series Accession`[1:20],
+   odir = "geo_cache"
+ )
Downloading 20 files
Download status: 20 done; 0 in progress. Total size: 95.55 Kb (100%)... done! 

> # 3. Start AI chat session
> chat <- geo_chat(meta, 
+                  provider = ellmer::chat_openai_compatible,
+                  base_url = "https://open.cherryin.cc/v1",
+                  model = "openai/gpt-4.1-mini", 
+                  system_prompt = "Be terse")
> # 4. Use AI to analyze and filter
> chat$chat("
+   Identify studies that:
+   1. Use scRNA-seq
+   2. Include treatment and control samples
+   
+   List the accession numbers that match these criteria.
+ ")
The studies that use scRNA-seq and include treatment and control samples are:

- GSE274638: "MYOD1L122R induces cancer stem cell pathways..." - single-cell RNA sequencing of 
mutant and wildtype PDX samples (treatment/control model)
- GSE253213: "Plastic state transitions and developmental potential predict response to Notch 
inhibition" - single-cell transcriptomics on treated and untreated T-ALL PDX models
- GSE306473: "Phase II Trial of Ixazomib combined with Gemcitabine and Doxorubicin in Patients 
with SMARCB1-Deficient Renal Medullary Carcinoma" - scRNAseq on samples from patients treated in 
the trial
- GSE299651: "Pooled single-cell screening in colorectal cancer identifies transcriptional 
modules..." - scRNA-seq of oncogene transfected cell lines, control and treatment-like conditions

These accessions meet both criteria of using scRNA-seq and having treatment and control samples.
> # 5. Get code to extract the filtered results
> chat$chat("
+   Write R code using dplyr to filter the metadata `meta`
+   for the studies you identified.
+ ")
```r
library(dplyr)

filtered_studies <- meta %>%
  filter(Series_geo_accession %in% c("GSE274638", "GSE253213", "GSE306473", "GSE299651"))
```
~~~

# Tips and Best Practices

1. **Be specific**: The more specific your questions, the better the AI can help.

2. **Ask for code**: The AI can generate R code for filtering, visualization, and analysis.

3. **Iterate**: Build on previous responses - the chat maintains conversation history.

4. **Verify results**: Always verify AI suggestions against your actual data.

5. **Use appropriate models**: For complex analyses, consider using more capable models like GPT-4.

# Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
