---
title: "AI-Powered GEO Metadata Exploration with geo_chat()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{AI-Powered GEO Metadata Exploration with geo_chat()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

The `geo_chat()` function enables interactive, AI-powered exploration of GEO
(Gene Expression Omnibus) metadata. By leveraging the
[ellmer](https://ellmer.tidyverse.org/) package, you can use natural language
to query, filter, and understand your GEO datasets.

## Prerequisites

Before using `geo_chat()`, you need to:

1. Install the `ellmer` package:
```{r install-ellmer}
install.packages("ellmer")
```

2. Configure an API key for your preferred AI provider. For example, for OpenAI:
```{r set-api-key}
# Set in your .Renviron file or directly:
Sys.setenv(OPENAI_API_KEY = "your-api-key-here")
```

Supported providers include:
- **OpenAI** (`ellmer::chat_openai`): Set `OPENAI_API_KEY`
- **Anthropic Claude** (`ellmer::chat_anthropic`): Set `ANTHROPIC_API_KEY`
- **Google Gemini** (`ellmer::chat_google_gemini`): Set `GOOGLE_API_KEY`
- And many more! See the [ellmer documentation](https://ellmer.tidyverse.org/) for the full list.

# Basic Usage

## Step 1: Get GEO Metadata

First, use geokit functions to retrieve metadata:

```{r setup}
library(geokit)

# Search for GEO Series related to your research topic
records <- geo_search("single cell RNA-seq[ALL] AND lung cancer[ALL] AND GSE[ETYP]")

# Get detailed metadata for the top results
meta <- geo_meta(records$`Series Accession`[1:10], odir = tempdir())
```

## Step 2: Start an AI Chat Session

Create a chat session with your metadata:

```{r create-chat}
chat <- geo_chat(meta, provider = ellmer::chat_openai)
```

## Step 3: Ask Questions!

Now you can interact with your metadata using natural language:
```{r chat-examples}
# Ask about the datasets
chat$chat("What types of studies are in this metadata?")

# Filter studies
chat$chat("Which studies have more than 50 samples?")

# Get R code suggestions
chat$chat("Write R code to filter studies that mention 'tumor' in the summary")

# Understand specific datasets
chat$chat("Tell me more about the study with the most samples")
```

# Interactive Console Mode

For a more immersive experience, use the interactive console:

```{r live-console}
ellmer::live_console(chat)
```

This opens an interactive chat interface where you can have a back-and-forth
conversation with the AI about your GEO metadata.

# Using Different AI Providers

## OpenAI (GPT-4, GPT-3.5)

```{r openai}
chat <- geo_chat(meta, provider = ellmer::chat_openai, model = "gpt-4o")
```

## Anthropic Claude

```{r anthropic}
chat <- geo_chat(meta, provider = ellmer::chat_anthropic)
```

## Google Gemini

```{r gemini}
chat <- geo_chat(meta, provider = ellmer::chat_google_gemini)
```

# Customizing the AI Behavior

You can add custom instructions to guide the AI:

```{r custom-prompt}
chat <- geo_chat(
  meta,
  provider = ellmer::chat_openai,
  system_prompt = "
    Focus on identifying single-cell RNA-seq studies.
    When suggesting R code, use tidyverse syntax.
    Always explain your reasoning step by step.
  "
)
```

# Controlling Context Size

For large metadata tables, you can control how much data is sent to the AI:

```{r max-rows}
# Only include first 50 rows in context
chat <- geo_chat(meta, provider = ellmer::chat_openai, max_rows = 50)

# Disable column statistics summary
chat <- geo_chat(meta, provider = ellmer::chat_openai, include_summary = FALSE)
```

# Example Workflow: Finding Single-Cell Cancer Studies

Here's a complete workflow for finding relevant single-cell cancer studies:

```{r workflow}
library(geokit)

# 1. Search for relevant studies
cancer_gse <- geo_search(
  "single cell[ALL] AND cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
)
cat("Found", nrow(cancer_gse), "potential studies\n")

# 2. Get detailed metadata
meta <- geo_meta(
  cancer_gse$`Series Accession`[1:20],
  odir = "geo_cache"
)

# 3. Start AI chat session
chat <- geo_chat(meta, provider = ellmer::chat_openai)

# 4. Use AI to analyze and filter
chat$chat("
  Identify studies that:
  1. Use 10x Genomics technology
  2. Have at least 10,000 cells
  3. Include treatment and control samples
  
  List the accession numbers that match these criteria.
")

# 5. Get code to extract the filtered results
chat$chat("
  Write R code using dplyr to filter the metadata 
  for the studies you identified.
")
```

# Tips and Best Practices

1. **Be specific**: The more specific your questions, the better the AI can help.

2. **Ask for code**: The AI can generate R code for filtering, visualization, and analysis.

3. **Iterate**: Build on previous responses - the chat maintains conversation history.

4. **Verify results**: Always verify AI suggestions against your actual data.

5. **Use appropriate models**: For complex analyses, consider using more capable models like GPT-4.

# Session Information

```{r session-info, eval=TRUE}
sessionInfo()
```
