---
title: "Build Your Own GEOmetadb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Build Your Own GEOmetadb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

If you are conducting bioinformatics research, especially when dealing with GEO
(Gene Expression Omnibus) data, you may have used the
[GEOmetadb](https://github.com/zhujack/GEOmetadb) package. While it was once a
great tool, unfortunately, it has not been updated for years. Additionally, the
`GEOmetadb` package is static, relying on the developer to manually update it,
which often results in incomplete or outdated data. 

**Why Build Your Own GEOmetadb?**

 - **Focus on Your Research Area**: For example, diabetes, liver cancer,
 urothelial carcinoma, etc. You can aggregate and annotate all relevant Series,
 making it easier for subsequent bulk downloads, filtering, and reproducible
 analysis. 
 - **Flexible Offline Querying**: GEO's web search is not flexible enough. After
 building your own database with `geokit`, you can use familiar tools like
 regular expressions or database queries (e.g., grep, SQLite) to freely search
 and meet various needs. 
 - **Dynamic Updates and Incremental Sync**: Automate scripts to regularly
 update the database, ensuring that the data is always up-to-date and avoiding
 the hassle of manual updates. 
 - **Efficient Data Access**: After storing the data offline, query speeds are
 significantly improved, especially when handling large GEO Series datasets,
 saving both time and bandwidth. 

`geokit` offers an efficient and straightforward method, with core operations
implemented in `rust`, allowing you to build your own metadata database in just
a few minutes. For example, processing 654 records takes around **30.5 seconds**
if the data is already downloaded.

 - Search the GEO database using **NCBI Eutils** and extract relevant metadata.
 - Fetch the relevant **metadata** from the GEO database and save it locally to
 build your own `GEOmetadb`. 
 - Use R's regular expressions, filtering features, or tools like Excel and
 SQLite to quickly search and analyze Series (GSE) and sample information.

# Search and Filter Single-Cell Studies of Urothelial Carcinoma

```{r setup}
library(geokit)
```

## 1. Search for Related Series by Keywords (e.g., Bladder/Urothelial Cancer)

Based on the **NCBI Eutils** query, more details can be found here: [Querying
GEO DataSets](https://www.ncbi.nlm.nih.gov/geo/info/qqtutorial.html)

```{r}
uc_gse <- list(
    geo_search(
        "bladder cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
    ),
    geo_search(
        "urothelial cancer[ALL] AND Homo sapiens[ORGN] AND GSE[ETYP]"
    )
)
uc_gse <- unique(dplyr::bind_rows(uc_gse))
```

## 2. Extract Sample Count from the "Contains" Field

```{r}
uc_gse$number_of_samples <- stringr::str_match(
  uc_gse$Contains, "(\\d+) Samples?"
)[, 2L, drop = TRUE]

# Quick Statistics
max_samples <- max(as.numeric(uc_gse$number_of_samples), na.rm = TRUE)
median_samples <- median(as.numeric(uc_gse$number_of_samples), na.rm = TRUE)
top_series <- dplyr::slice_max(uc_gse, as.numeric(number_of_samples))
```

## 3. Fetch Series Metadata (Parallel Processing and Batch Saving to Local Directory)

```{r, eval=FALSE}
uc_gse_meta <- geo_meta(uc_gse[["Series Accession"]],
  odir = "gse_urothelial_cancer"
)
```

## 4. Filter Possible Single-Cell Studies (Search for Keywords in Summary/Title/Design)

```{r, eval=FALSE}
uc_gse_sc <- dplyr::filter(
  uc_gse_meta,
  dplyr::if_any(
    c(Series_summary, Series_title, Series_overall_design),
    grepl,
    pattern = "single[- ]cell|scRNA", ignore.case = TRUE
  )
) |>
  dplyr::mutate(
    number_of_samples = lengths(
      strsplit(Series_sample_id, "; ", fixed = TRUE)
    )
  )

# Output Results and Statistics
dplyr::slice_max(uc_gse_sc, number_of_samples)$Series_geo_accession
max(uc_gse_sc$number_of_samples, na.rm = TRUE)
median(uc_gse_sc$number_of_samples, na.rm = TRUE)
writexl::write_xlsx(uc_gse_sc, "uc_gse_sc.xlsx")
```

By following these steps, you have successfully created your own `GEOmetadb`!


# Session Information
```{r}
sessionInfo()
```
